---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r, results="hide"}
library(dplyr)
library(lattice)
```


## Loading and preprocessing the data

```{r, echo=TRUE}

# download zip file
# to force a download uncomment the if statement below

datasource <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if (!file.exists("activity.zip")) 
{
	download.file(datasource, destfile = "activity.zip", method = "curl")
}

# unzip the data

datafile <- "activity.csv"
if (!file.exists(datafile)) 
{
	unzip("activity.zip")
}

# load data

activity <- read.csv(datafile, header = TRUE)
#nrow(activity)
head(activity)
summary(activity)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r, echo=TRUE}

total_steps_per_day <- activity %>% 
                       group_by(date) %>% 
                       summarize(total_steps = sum(steps, na.rm = TRUE))

hist(total_steps_per_day$total_steps,
     main="Histogram: Total Steps per Day",
     xlab="Total Steps per Day",
     ylab="Frequency",
     col="blue")

mean_total_steps <- mean(total_steps_per_day$total_steps)
median_total_steps <- median(total_steps_per_day$total_steps)
```

==> Median number of steps taken each day: `r median_total_steps`.  
==> Mean number of steps taken each day: `r mean_total_steps`.  
  

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}

daily_pattern <- activity %>% 
                 group_by(interval) %>% 
                 summarize(average_steps = mean(steps, na.rm = TRUE))

daily_pattern
#daily_pattern$average_steps

plot(daily_pattern, 
     type = "l", 
     main = "Average Daily Activity Pattern",
     col = "blue")

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}

max_index <- which.max(daily_pattern$average_steps)
#max_index
#daily_pattern[max_index,1]

```

==> The maximum average number of steps occur at interval index `r max_index` (`r daily_pattern[max_index,1]`)  
  

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs):
```{r, excho=TRUE}

# calculate total number of missing values in the dataset

total_missing_values <- sum(is.na(activity))
total_missing_values
```
==> The total number of missing values in the dataset is `r total_missing_values`.  
  
2. Devise a strategy for filling in all of the missing values in the dataset.

==> The missing values will be filled in with the (rounded) mean for the 5-minute interval

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  

```{r, echo=TRUE}

# create a new dataset equal to the original

activity.imputed <- activity

# run through, substituting all NA values with the mean

for (i in 1:nrow(activity.imputed)) 
{
    if (is.na(activity.imputed$steps[i]))
    {
       avg_steps <- daily_pattern[daily_pattern$interval == activity.imputed$interval[i],"average_steps"]
       activity.imputed$steps[i] <- as.integer(round(avg_steps))  
    }
}

# debug lines
# head(activity.imputed,20)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r, echo=TRUE}

imputed_steps_per_day <- activity.imputed %>% 
                         group_by(date) %>% 
                         summarize(imputed_steps = sum(steps, na.rm = TRUE))

hist(imputed_steps_per_day$imputed_steps,
     main="Histogram: Imputed Steps per Day",
     xlab="Imputed Steps per Day",
     ylab="Frequency",
     col="blue")

mean_imputed_steps <- mean(imputed_steps_per_day$imputed_steps)
median_imputed_steps <- median(imputed_steps_per_day$imputed_steps)

```

==> (Imputed) Median number of steps taken each day: `r median_imputed_steps`.  
==> (Imputed) Mean number of steps taken each day: `r as.integer(mean_imputed_steps)`. 

#### Observations
==> Imputing missing data with average values will have a normalising effect on the histogram.  
==> Imputing missing data with average values also brings the mean and median much closer to each other.


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}

# add qualifyers to imputed dataset indicating week or weekend days

activity.imputed <- mutate(activity.imputed, day = weekdays(as.Date(date)))
activity.imputed <- mutate(activity.imputed, daytype = as.factor(ifelse(day == "Saturday" | day == "Sunday", "Weekend", "Weekday")))
head(activity.imputed)

# create daily pattern to include weekend qualifier as key

daily_pattern_by_daytype <- activity.imputed %>% 
                 group_by(daytype,interval) %>% 
                 summarize(average_steps = mean(steps))
head(daily_pattern_by_daytype)
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE}

# plot the data split by weekend and weekday graphs

xyplot(average_steps ~ interval | daytype, 
       data = daily_pattern_by_daytype, 
       layout = c(1,2), 
       type = "l",
	    main = "Average Steps by 5-minute Interval for Weekend vs Weekdays",
	    xlab = "Interval",
	    ylab = "Average Steps",
	    col = "blue")
```

#### Observation

==> Weekdays seems to have a higher activity level early in the morning, whereafter it flattens out.  
==> Weekends have a higher activity level throughout the rest of the day  

